---
title: "data-access"
format: html
---

# R programmatic data access

## Cache/Check
See if a data file exists. If it does, use it. If it does not, dowload it to a defined data directory.

```{r}
# Set paths
yolo_catch_url <- "https://pasta.lternet.edu/package/data/eml/edi/233/2/015e494911cf35c90089ced5a3127334"
data_dir <- "data"
yolo_catch_file <- file.path(data_dir, "yolo_catch.csv")

# Download data if not already cached
if (!file.exists(yolo_catch_file)) {
  message("Downloading file...")
  dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
  download.file(
    url = yolo_catch_url,
    destfile = yolo_catch_file
  )
  message("Saved to: ", yolo_catch_file)
} else {
  message("Using cached file.")
}

# Load the (locally cached) data
yolo_catch <- readr::read_csv(yolo_catch_file, guess_max = 100000)
head(yolo_catch)
```

### Notes on rds vs csv
If you’re caching an R data frame (rather than a raw file), prefer RDS format over CSV:

Saves the data frame with its types preserved
saveRDS(my_data, "data/my_data.rds")

Restores the data frame exactly, keeping dates as dates and factors as factors
my_data <- readRDS("data/my_data.rds")

CSV loses column type information (dates become character strings, and you have to re-specify types every time you load). RDS is faster to read and write, and perfectly reconstructs the original object.

## Practice API data access with USGS water monitoring data
```{r}
# Get all stream gauge stations in California
ca_streams <- read_waterdata_monitoring_location(
  state_name = "California",
  site_type = "Stream"
)
```

### Further filter data to a smaller spatial area
```{r}
# Build a bounding box for a region of interest, then filter spatially
delta_bbox <- sf::st_bbox(c(xmin = -122.1, ymin = 37.8,
                            xmax = -121.2, ymax = 38.5),
                          crs = sf::st_crs(4326))
delta_sites <- sf::st_filter(ca_streams, sf::st_as_sfc(delta_bbox))

# How many stations?
nrow(delta_sites)

# Preview results (geometry column is retained automatically)
delta_sites |>
  select(monitoring_location_id, monitoring_location_name) |>
  head(10)
```

### map the Freeport station
```{r}
options(tigris_use_cache = TRUE)

# Station of interest (already an sf object from delta_sites)
freeport <- delta_sites |>
  filter(monitoring_location_id == "USGS-11447650")

# Compute map extent: ±0.35° around the station
lon <- as.numeric(sf::st_coordinates(freeport)[1, "X"])
lat <- as.numeric(sf::st_coordinates(freeport)[1, "Y"])
buf <- 0.35
map_xlim <- c(lon - buf, lon + buf)
map_ylim <- c(lat - buf, lat + buf)

# Find counties intersecting the visual extent
vis_rect <- sf::st_sfc(
  sf::st_polygon(list(matrix(c(
    map_xlim[1], map_ylim[1], map_xlim[2], map_ylim[1],
    map_xlim[2], map_ylim[2], map_xlim[1], map_ylim[2],
    map_xlim[1], map_ylim[1]
  ), ncol = 2, byrow = TRUE))),
  crs = sf::st_crs(4326)
)
all_ca <- counties(state = "CA", year = 2022, progress_bar = FALSE)
area_counties <- sf::st_filter(all_ca,
    sf::st_transform(vis_rect, sf::st_crs(all_ca)))

# Water-body polygons for those counties
area_water_sf <- map(area_counties$NAME,
    ~area_water("CA", county = .x, year = 2022)) |>
  bind_rows()

# Extract Freeport station coordinates for the text label (avoids
# geom_sf_text CRS warning)
site_coords <- as.data.frame(sf::st_coordinates(freeport))

ggplot() +
  geom_sf(data = area_counties, fill = "#f5f0e8",
          color = "gray60", linewidth = 0.3) +
  geom_sf(data = area_water_sf, fill = "#a6cee3",
          color = "#6baed6", linewidth = 0.1) +
  geom_sf(data = freeport, color = "#e31a1c",
          size = 4, shape = 21, fill = "#e31a1c") +
  geom_text(data = site_coords,
            aes(x = X, y = Y, label = "Freeport\n(11447650)"),
            nudge_x = 0.05, hjust = 0, size = 2.8) +
  coord_sf(xlim = map_xlim, ylim = map_ylim, expand = FALSE) +
  labs(title = "Sacramento River at Freeport (USGS 11447650)",
       x = NULL,
       y = NULL) +
  theme_light() +
  theme(panel.background = element_rect(fill = "#e8f4f8"))
```

### explore data availability

```{r}
# What data does Freeport have?
freeport_avail <- read_waterdata_combined_meta(
  monitoring_location_id = "USGS-11447650"
)

# Which parameters are available, and for how long?
freeport_avail |>
  select(parameter_code, parameter_name, statistic_id, begin, end) |>
  arrange(parameter_code, statistic_id)
```

```{r}
freeport_avail |>
  filter(parameter_code == "00010") |>
  count(statistic_id)
```


Although you can manually look up names and descriptions of all possible statistic codes in a browser, you can also retrieve this information directly via the package:
```{r}
stat_codes <- read_waterdata_metadata("statistic-codes")

# Show statistic names along with other station availability data
freeport_avail %>%
  filter(parameter_code == "00010") %>%
  select(parameter_code, parameter_name, statistic_id, begin, end) %>%
  left_join(stat_codes, by = join_by(statistic_id == statistic_code)) %>%
  select(-statistic_description) %>%
  relocate(statistic_name, .after = statistic_id) %>%
  arrange(parameter_code, statistic_id)
```

read_waterdata_daily() retrieves daily values. We pass parameter_code = "00010" to select water temperature and statistic_id = "00003" to select the daily mean, and wrap the call in our caching pattern:

```{r}
freeport_temp_file <- "data/freeport_temp.rds"

if (!file.exists(freeport_temp_file)) {
  message("Fetching daily temperature from USGS...")
  freeport_temp <- read_waterdata_daily(
    monitoring_location_id = "USGS-11447650",  # note "USGS-" prefix
    parameter_code = "00010",  # water temperature
    statistic_id = "00003",  # daily mean
    time = c("2020-01-01", "2023-12-31")
  )
  dir.create("data", showWarnings = FALSE)
  saveRDS(freeport_temp, freeport_temp_file)
} else {
  freeport_temp <- readRDS(freeport_temp_file)
}

glimpse(freeport_temp)
```

### visualize the data

```{r}
ggplot(freeport_temp, aes(x = time, y = value)) +
  geom_line(color = "#e34a33", alpha = 0.6, linewidth = 0.3) +
  labs(
    x = NULL,
    y = "Temperature (°C)",
    title = "Daily Mean Water Temperature at the Sacramento River near Freeport",
    subtitle = "USGS Station 11447650 (2020-2023)"
  ) +
  theme_light()
```

```{r}
freeport_temp |>
  mutate(month = month(time, label = TRUE)) |>
  ggplot(aes(x = month, y = value)) +
  geom_boxplot(fill = "#8abcfd", outlier.alpha = 0.2, outlier.size = 0.5) +
  labs(
    x = "Month",
    y = "Temperature (°C)",
    title = "Seasonal Temperature Distribution",
    subtitle = "Sacramento River near Freeport (2020-2023)"
  ) +
  theme_light()
```

### fetching multiple parameters at once

```{r}
multi <- read_waterdata_daily(
  monitoring_location_id = "USGS-11447650",
  parameter_code = c("00010", "00060", "80154"),  # temp, discharge, susp sedmnt conc
  statistic_id = "00003",  # daily mean for all three
  time = c("2020-01-01", "2023-12-31")
)

# Plot time series faceted by parameter
multi %>%
  ggplot(aes(x = time, y = value, col = parameter_code)) +
  geom_line() +
  facet_wrap(~parameter_code, scales = "free_y", ncol = 1)
```

### other potentially useful data sources that can be pulled in
dataRetrieval also connects to the Water Quality Portal, which aggregates data from USGS, EPA, and over 400 state and local agencies. Use readWQPdata() to query this broader dataset:
```{r}
# Water quality data from all agencies at a USGS site
wq <- readWQPdata(
  siteid = "USGS-11447650",
  characteristicName = "Turbidity"
)
```



### Activity: Retrieve and visualize water data
Working individually or in pairs (10 min), choose one of the following:

#### Option A: Different station: 
Retrieve daily mean water temperature (parameter_code = "00010", statistic_id = "00003") for one of these other nearby stations and compare the seasonal pattern with Freeport. Is the annual cycle the same shape? Earlier or later in the year?

San Joaquin River at Vernalis: USGS-11303500
Cache Slough near Ryer Island: USGS-11455385
Sacramento River at Rio Vista: USGS-11455420

```{r}
# What data does Rio Vista have?
riovista_avail <- read_waterdata_combined_meta(
  monitoring_location_id = "USGS-11455420"
)

# Which parameters are available, and for how long?
riovista_avail |>
  select(parameter_code, parameter_name, statistic_id, begin, end) |>
  arrange(parameter_code, statistic_id)

riovista_avail |>
  filter(parameter_code == "00010") |>
  count(statistic_id)

```

```{r}
riovista_temp_file <- "data/riovista_temp.rds"

if (!file.exists(riovista_temp_file)) {
  message("Fetching daily temperature from USGS...")
  riovista_temp <- read_waterdata_daily(
    monitoring_location_id = "USGS-11455420",  # note "USGS-" prefix
    parameter_code = "00010",  # water temperature
    statistic_id = "00003",  # daily mean
    time = c("2020-01-01", "2023-12-31")
  )
  dir.create("data", showWarnings = FALSE)
  saveRDS(riovista_temp, riovista_temp_file)
} else {
  riovista_temp <- readRDS(riovista_temp_file)
}

glimpse(riovista_temp)
```

#### Option B: Different statistic: 
At Freeport (USGS-11447650), retrieve the daily maximum (statistic_id = "00001") and daily minimum ("00002") temperature in addition to the mean. Plot all three on one time series. How wide is the daily range? Does it vary by season?

#### Option C: Temperature vs. discharge: 
Retrieve both daily mean temperature (00010) and daily mean suspended sediment concentration (80154) at Freeport. Make a scatter plot. Is there a clear relationship? What might you look at next?

Include the cache pattern in your code so it only downloads once!



