---
title: "Working with Vector Spatial Data in R"
author: "ME!"
format: html
---

## Setup
### Download data
```{r}
out_dir <- here::here('data/ak_regions')

if(!dir.exists(out_dir)) {
    knb_url <- "https://dev.nceas.ucsb.edu/knb/d1/mn/v2/object/urn%3Auuid%3Aaceaecb2-1ce0-4d41-a839-d3607d32bb58"

    download.file(url = knb_url, 
                  destfile = 'shapefile_demo_data.zip', 
                  mode = 'wb')

    unzip('shapefile_demo_data.zip', exdir = out_dir)

    file.remove('shapefile_demo_data.zip')
}
```

### Load packages
```{r}
#| message: false
#| warning: false

### Data wrangling packages:
library(readr)
library(dplyr)
library(tidyr)
library(here)

### Simple Features vector spatial data:
library(sf)

### Data visualization packages:
library(ggplot2)
library(leaflet)
library(scales)
library(ggspatial)
```


## Data Exploration
### Read in Data
```{r}
ak_rgns_sf <- read_sf(here("data/ak_regions/ak_regions_simp.shp"))

#sf means simple features to know it is not just a data frame but, in this case, a shapefile

```

### Transform projection (ESPG code) for better visualization of Alaska

```{r}
ak_rgns_3338_sf <- ak_rgns_sf %>%
    st_transform(crs = 3338)

# st_crs(ak_rgns_3338_sf) ### run in console to check
```


```{r}
#| eval: false

### Quick Views of data

# for more information about your file, can run these in the console
# class(ak_rgns_3338_sf)
# can see that there is a data.frame class which allows us to play with it and analyze/manipulate using tidyverse
# head(ak_rgns_3338_sf)
# glimpse(ak_rgns_3338_sf)
```

### Spatial Joins
Suppose you ask the question: how many people live in each of these regions in Alaska?

#### download the population data

```{r}
out_file <- here::here('data/alaska_population.csv')

if(!file.exists(out_file)) {
    pop_data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3A7fc6f6db-c5ea-426a-a743-1f2edafb43b8"

    download.file(url = pop_data_url, 
                  destfile = out_file, 
                  mode = 'w')
}
```

#### read in population data

```{r}
pop_df <- read_csv(here("data/alaska_population.csv")) %>%
    filter(Year == 2015) %>%
    select(city, lat, lng, population = total) %>%
    drop_na()

# now convert pop_df from dataframe to a sf
# note that because it is in lat/long, need to read in as crs 4326 which is the underlying lat long projection. Later, can transform to other projection.

pop_4326_sf <- st_as_sf(pop_df,
                        coords = c('lng', 'lat'),
                        crs = 4326,
                        remove = FALSE)
```

#### Join population and region data

```{r}
# first transform population data to same CRS as the AK region data, 3338
pop_3338_sf <- st_transform(pop_4326_sf, 
                            crs = 3338)

# then we can join the two
pop_joined_sf <- st_join(pop_3338_sf, 
                         ak_rgns_3338_sf, 
                         join = st_within)

ak_rgns_joined_sf <- st_join(ak_rgns_3338_sf,
                             pop_3338_sf,
                             join = st_within)
```


## Analysis
### calculate total population by region

```{r}
pop_rgn_df <- pop_joined_sf %>%
    st_drop_geometry() %>%
    group_by(region) %>%
    summarize(total_pop = sum(population))
```



## Visualization

### Map population by region

```{r}
# join total population by region with ak shapefile to map population by region
pop_rgn_3338_sf <- left_join(ak_rgns_3338_sf, 
                             pop_rgn_df, 
                             by = "region")

### plot to check
plot(pop_rgn_3338_sf["total_pop"])
```

### Dissolve geometries to map by management area

```{r}
pop_mgmt_3338_sf <- pop_rgn_3338_sf %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop))

plot(pop_mgmt_3338_sf["total_pop"])
```

```{r}
#| eval: false
# save to your files
write_sf(pop_rgn_3338_sf, here("data/ak_regions_population.gpkg"))
```

### Mapping using ggplot2
```{r}
ggplot(pop_rgn_3338_sf) +
    geom_sf(aes(fill = total_pop)) +
    labs(fill = "Total Population") +
    scale_fill_continuous(low = "khaki", high =  "firebrick",
                          labels = comma, name = "Population") +
    theme_bw()
```

#### Add rivers, cities data to map
##### Read in rivers data
```{r}
rivers_3338_sf <- read_sf(here("data/ak_regions/ak_rivers_simp.shp"))
```

#### Map it!

```{r}
# note: order matters here. what you put first will be on the bottom.
ggplot() +
    geom_sf(data = pop_rgn_3338_sf, 
            mapping = aes(fill = total_pop)) +
    geom_sf(data = rivers_3338_sf,
            aes(linewidth = StrOrder),
            color = 'blue', alpha = .5) +
    scale_linewidth(range = c(0.05, 0.5),
                    guide = "none") +
    geom_sf(data = pop_3338_sf, size = 0.5, color = 'grey20') +
    labs(title = "Total Population by Alaska Region",
         fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw() 
```



