---
title: "Function practice"
author: "BrookeE"
format: html
---

## Setup
### Load packages

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr) ### for iteration using the map() family of functions

library(palmerpenguins)
```

### simple function anatomy
```{r}
airtemps <- c(212, 30.3, 78, 32)
celsius1 <- (airtemps[1] - 32) * 5/9
celsius2 <- (airtemps[2] - 32) * 5/9
celsius3 <- (airtemps[3] - 33) * 5/9   ### repeating calculations can inadvertently introduce errors. Better to use a function.

### write function that converts fahrenheit temp to celsius
convert_f_to_c <- function(fahr) {
    celsius <- (fahr - 32) * 5/9
    return(celsius)
}
celsius1a <- convert_f_to_c(fahr = airtemps, scale = 'c')
```

### a little more complex
```{r}
### now create a function that will convert either way, user describes F or C
convert_temp <- function(t, scale = 'f') {  ### scale = 'f' tells function that if user doesn't include scale, assume it is f
    scale = tolower(substr(scale, 1,1))   ### this allows fahrenheit or celsius to be entered in function and it will extract the first letter to work with the function

    if(!scale %in% c('c','f')) {  
        stop('Scale must be either "c" or "f"!')  ### this creates a custom error for users to see if they don't write c or f
    }
        if(scale == 'c') {
        result <- (t * 9/5) + 32
    } else {
        result <- (t-32) * 5/9
    }
    return(result)
}

convert_temp(t = airtemps, scale = 'celsius')
```

#### thinking about what returns and in what format
```{r}
### for more complex returns, can choose how it is returned.
### like in a data frame:
convert_f_to_c_k <- function(f) {
  c <- (f - 32) * 5/9
  k  <- c + 273.15
  out_df <- data.frame(fahr = f, celsius = c, kelvin = k)
  return(out_df)
}

temps_df <- convert_f_to_c_k(seq(-100, 100, 50))
temps_df
```


```{r}
### or you may want to return a list
convert_f_to_c_k <- function(f) {
  c <- (f - 32) * 5/9
  k  <- c + 273.15
  out_list <- list(fahr = f, celsius = c, kelvin = k)
  return(out_list)   ### even without writing this, R will return the last line in the function (in this case, out_list)
                     ### and everything after the return line will be ignored
}

temps_list <- convert_f_to_c_k(seq(-100, 100, 50))
temps_list
```


### Using purrr and interating functions across vectors or df

```{r}
library(palmerpenguins)
data(penguins)
spp_vec <- unique(penguins$species)

calc_bill_model <- function(spp, df) {
    spp_df <- df %>%
        filter(species == spp)
    bill_mdl <- lm(bill_length_mm ~ bill_depth_mm, data = spp_df)
    return(bill_mdl)
}
map_results <- purrr::map(.x = spp_vec, .f = calc_bill_model, df = penguins)
broom::tidy(map_results[[3]])
# you can do the same exact thing as above (purrr) with base r code:
lapply_results <- base::lapply(X = spp_vec, FUN = calc_bill_model, df = penguins)
broom::tidy(lapply_results[[3]])
```

### Functions are great for creating plots and plot themes too

```{r}
custom_theme <- function(base_size = 9) {
  theme(
    text             = element_text(family = "serif",
                                    color  = "slateblue4",
                                    size   = base_size),
    plot.title       = element_text(size   = rel(1.25),
                                    hjust  = 0.5,
                                    face   = "bold"),
    panel.background = element_rect(color = 'slateblue3',
                                    fill  = 'azure'),
    panel.grid.major = element_line(color    = "slateblue1",
                                    linewidth = 0.25),
    legend.position  = c(.9, .4),
    axis.ticks       = element_line(color = 'red')
  )
}

#### and you can go further to put the plot creation in a function
scatterplot <- function(df, point_size = 2, font_size = 9) {
  ggplot(data = df, mapping = aes(x = fahr, y = celsius, color = kelvin)) +
    geom_point(size = point_size) +
    scale_color_viridis_c() +
    custom_theme(font_size)
}

scatterplot(temps_df, point_size = 3, font_size = 16) +
  labs(title = 'Temperature Conversions')
```

## Documenting Functions
Well-named functions are a start, but good documentation tells collaborators (and future you) what a function expects and what it returns. The example below uses the roxygen2 (document generation) package syntax and example to package your function with a help doc for R that will return with this useful information for users.

```{r}
#' Convert temperature from Fahrenheit to Celsius
#'
#' @param fahr Numeric value or vector in degrees Fahrenheit
#'
#' @returns Numeric value or vector in degrees Celsius
#' @export
#'
#' @examples
#' convert_f_to_c(32)
#' convert_f_to_c(c(32, 212, 72))
convert_f_to_c <- function(fahr) {
  celsius <- (fahr - 32) * 5/9
  return(celsius)
}
```

## Test your new knowledge
Write a function that takes salmon length, in inches, and returns the corresponding weight, in pounds. The simple model is 
; for Chinook salmon, let’s use W=aL^b and (for inches to pounds). Give it a name that indicates the purpose of the function. For testing purposes, a 30” salmon should be about 11 lb; a 42” salmon is about 27 lb; and a 60” is about 82 lb.

```{r}
convert_l_to_wt <- function(L) {
    a <- 0.00057
    b <- 2.9  ### inches to pounds
    salmon_length <- a * (L^b)
    return(salmon_length)
}

salmon_weight <- c(30, 42, 60)
convert_l_to_wt(L=salmon_weight)
```
